The Impact of Alternation

### 2017-05-03 draft.tex
ok

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
words over $\vec{x}$. This definition extends the classical notion of
words from a finite alphabet with a possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be
implemented in linear time. This is not very surprising, since linear
time intersection and complementation are possible for alternating
automata over finite alphabets \cite{ChandraKozenStockmeyer81}.

\section{Antichains for Emptiness}

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
antichain\footnote{Given a partial order $(D,\preceq)$ an antichain is
  a set $A \subseteq D$ such that $a \not\preceq b$ for any $a,b \in
  A$.} w.r.t. the entailment partial order. This is because, for any
$u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models \Post{\A}(\iota,v)$
and $\Accept{\A}(uw)$ is satisfiable, then $\Accept{\A}(vw)$ is
satisfiable as well, and there is no need to store
$\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be proved
using $v$ alone. However, in practice, the semi-algorithm using
antichains diverges in a large number of cases, because, in many
interpretations, there exist infinite antichains\footnote{For
  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq
  0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u) \mid u \in U}$, where
$U$ is an infinite subset of $\Sigma^*$.

A \emph{safety invariant} for $\A$ is a function $I : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, the following
hold: \begin{compactitem}
\item if $\beta \models \iota$ then $\nu \in I(\beta)$ iff
  $\nu(x)=\mathbf{0}$ for all $x \in \vec{x}$,
%
\item for each $u \in \Sigma^*$, if $\beta \cup \nu \models
  \Post{\A}(\iota,u)$ then $\nu \in I(\beta)$, and
%
\item 
\end{compactitem}

>>>>

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
words over $\vec{x}$.

This definition extends the classical notion of words from a finite
alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
when $\Data^\I$ is sufficiently large or infinite, we can map the
elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
special variable to encode the input events. However, keeping $\Sigma$
explicit in the following simplifies several technical points below,
without cluttering the presentation.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This is not very surprising, since linear time
intersection and complementation are also possible for alternating
automata over finite alphabets \cite{ChandraKozenStockmeyer81}.

\section{Antichains and Abstraction Refinement for Emptiness}

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need to
store $\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be
proved using $v$ alone. However, in practice, the semi-algorithm using
antichains diverges in a large number of cases, because, in many
interpretations, there exist infinite antichains\footnote{For
  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge
  x \teq 0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u_i) \mid
  u_1,u_2, \ldots \in \Sigma^*}$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactitem}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item $\beta \cup \nu \models \Post{\A}(\iota,u) \Rightarrow \nu \in
  \Inv(\beta)$, and
%
\item $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu \not\models
  \Accept{\A}(u)$.
\end{compactitem}
Intuitively, a safety invariant maps every boolean valuation into a
set of data valuations, that contains the initial configuration (in
which all data variables are assumed to be $\mathbf{0}\in\Data^\I$),
over-approximates the set of reachable valuations and excludes those
valuations that satisfying the acceptance condition.

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A)
= \emptyset$ iff $\A$ has a safety invariant. 
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define a function $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x}
    \mapsto \Data^\I}$ as: for each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is indeed a safety
  invariant is straightforward. \qed}

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post(\iota,u_i)$ defines
a safety invariant for $\A$. Although this condition can be
effectively checked using a decision procedure for the quantifier-free
first order theory $\theory(\sig,\I)$, of signature $\sig$ and
interpretation $\I$, there is no guarantee that this check will ever
succeed.
